Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2024-07-23 ms.qmd
Few updated sentences
Fixed sentence

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can; @Sanderson2023; @urban2023and]. Here we show that both acclimation responses ($lnRR_{Q_{10}}$) and increases in physiological rate variance ($lnVR_{Q_{10}}$) of ectotherms varied across habitats. Our results uncover an hitherto unrecognised dynamic where the benefits of partial acclimation are paralleled by increases in trait variance that may have impacts on how ectotherm populations across different habitats will be able to adapt to increased temperatures.

@@

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change. Assuming that changes in variation in physiological rates are underpinned by genetic variation, and that there is a genetic correlation with fitness, smaller increases in physiological variance could limit adaptation in terrestrial habitats more than aquatic habitats in the future [@urban2023and; @hoffmann2011climate]. For example, under climate change we expect an increase in variance in physiological rates of only ~13% in terrestrial habitats whereas for freshwater habitats we expect variation in physiological rates to increase by ~30%. There will obviously be limits to variance increases, and we predict that organisms closer to their upper thermal limits ($CT_{max}$) will have lower $lnVR_{Q_{10}}$ values compared to those farther away from $CT_{max}$. Some evidence points to possible differences across habitats in upper thermal limits already [@Gunderson2015-hz; @Pinsky2019-fn], making this a fruitful future question to explore.

>>>>

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can; @Sanderson2023; @urban2023and]. Here we show that both acclimation responses ($lnRR_{Q_{10}}$) and increases in physiological rate variance ($lnVR_{Q_{10}}$) of ectotherms varied across habitats. Our results uncover an hitherto unrecognised dynamic where the benefits of partial acclimation are paralleled by increases in trait variance that depend on habitat in ways that may have impacts on how ectotherm populations will be able to adapt to increased temperatures.

@@

Our results further highlight the potential vulnerability of terrestrial ectotherms to climate change. Assuming that changes in variation in physiological rates are underpinned by genetic variation, and that there is a genetic correlation with fitness, smaller increases in physiological variance could limit adaptation in terrestrial habitats more than aquatic habitats in the future [@urban2023and; @hoffmann2011climate]. For example, under climate change we expect an increase in variance in physiological rates of only ~13% in terrestrial habitats whereas for freshwater habitats we expect variation in physiological rates to increase by ~30%. Importantly, responses to selection will also depend on the magnitude and direction of genetic covariances with other traits, which need consideration. There will obviously be limits to variance increases, and we predict that organisms closer to their upper thermal limits ($CT_{max}$) will have lower $lnVR_{Q_{10}}$ values compared to those farther away from $CT_{max}$. Some evidence points to possible differences across habitats in upper thermal limits already [@Gunderson2015-hz; @Pinsky2019-fn], making this a fruitful future question to explore.
